---
title: "`r config$publication_name`"
pagetitle: DFM England summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: 
      - www/style.css
      - www/nhs-min.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<html lang="en">

```{r setup, include=FALSE}

# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<div class = "quartoheader", id = "fixed-header">
  <header class="nhsuk-header" role="banner">
    <div class="nhsuk-width-container nhsuk-header__container">
      <div class="nhsuk-header__logo nhsuk-header__logo--only">
        <a class="nhsuk-header__link" href="https://www.nhsbsa.nhs.uk/" aria-label="NHSBSA home" target="_blank">
          <img class="nhsuk-logo" src="www/logo-nhsbsa.svg"" name="NHSBSA logo" alt="NHS Business Services Authority">
        </a>
      </div>
    </div>
</div>

<!-- JS to handle hiding the header on scroll -->
<script>
  document.addEventListener("scroll", function() {
    var header = document.getElementById("fixed-header");
    var toc = document.getElementById("TOC");
    var BackTopButton = document.getElementById("backtop-button");
    var scrollPosition = window.scrollY || document.documentElement.scrollTop;
  if(scrollPosition > 88) {
    header.style.top = "-200px";
    toc.style.setProperty('margin-top', '24px', 'important');
    BackTopButton.style.setProperty('visibility', 'visible', 'important');
  } else
  {
    header.style.top = "0";
    toc.style.setProperty('margin-top', '140px', 'important');
    BackTopButton.style.setProperty('visibility', 'hidden', 'important');
  }
}
);
</script>

<!-- JS to handle back to top button -->
<script>
       function BackTop() {
        document.body.scrollTop = 0;
        document.documentElement.scrollTop = 0;
    }
</script>

<button onclick="BackTop()" id="backtop-button" style="opacity: 0.7; pointer-events: initial; visibility: visible;"><span aria-hidden="true">▲</span>Back to top</button>

<main>

# `r config$publication_sub_name` {.toc-ignore}

`r paste0("Published ", config$publication_date)`

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have added supporting summary tables containing data presented by calendar year and financial quarter. These tables contain the same geographical and demographic breakdowns as the financial year tables.<br><br>We have also updated our methodology for assigning identified patients to gender groups. You can find more information in Section 4 - About these statistics.<br><br>We are interested in any feedback about the publication, which you can send by using our ", htmltools::HTML(paste("<a href='",config$stats_survey_link,"'>Official Statistics feedback survey</a>")),"."), width = "100%")` 
::: 
::::

## Key findings

In England in 2023/24:

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>67 million items for dependency-forming medicines were prescribed.</b>", width = "100%")` 
::: 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>There was a 52% decrease in costs for dependency-forming medicines prescribed from 2015/16.</b>", width = "100%")` 
:::
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>7.1 million identified patients were prescribed dependency-forming medicines.</b>", width = "100%")` 
:::
::::

There were 67 million items for dependency-forming medicines prescribed, a 1% decrease from 2015/16.

The cost of dependency-forming medicines prescribed in England was £370 million. This was a 52% decrease from 2015/16 when the cost was £780 million.

Opioid drugs were the most prescribed dependency-forming medicines with 39 million items at a cost of £280 million. The total cost of opioid drugs has decreased by 34% since 2015/16.

There were 7.1 million identified patients that were prescribed dependency-forming medicines. This was a 13% decrease from 8.1 million identified patients in 2015/16.

The most common group to be prescribed dependency-forming medicines was female patients aged 60 to 64 with 420,000 identified patients.

Areas of greater deprivation had the highest number of identified patients who were being prescribed dependency-forming medication. 57% more patients received prescribing in the most deprived areas of the country compared to those in the least deprived.

---

## 1. Things you should know

### 1.1. Background {.toc-ignore}

This  publication was developed in response to the Public Health England (PHE) review into the dependence and withdrawal associated with some prescribed medicines.

Known as the [prescribed medicines review (PMR)](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/940255/PHE_PMR_report_Dec2020.pdf), it recommended an increase in the availability and use of data on the prescribing of medicines that can cause dependence.

This publication includes data on 5 categories of medicines overall:

*	Antidepressants
*	Opioid pain medicine
*	Gabapentinoids
*	Benzodiazepines
*	Z-drugs

Antidepressants are not included in the measures for volume, cost, or demographics. The [current National Institute for Health and Care Excellence (NICE) guidance](https://www.nice.org.uk/guidance/ng215) makes the distinction that antidepressants can cause withdrawal symptoms but are historically not dependency-forming. Additionally, the statistics for antidepressants can be found in the [Medicines Used in Mental Health publication](https://www.nhsbsa.nhs.uk/statistical-collections/medicines-used-mental-health-england). Antidepressants are included in the co-prescribing measures in this publication.

This publication aims to describe the prescribing of dependency-forming medicines in England that are dispensed in the community. This does not include data on medicines prescribed and dispensed in secondary care, prisons, or issued by a private prescriber.

These statistics detail:

* the total number of prescription items issued for dependency-forming medication
* the total cost of prescription items issued for these drugs
* the number of identified patients that have received prescribing for these drugs
* the number of identified patients receiving more than one dependency-forming medication
* demographic breakdowns of prescribing by age group and gender
* demographic breakdowns by gender
* demographic breakdowns by a measure of deprivation.

These statistics use the BNF therapeutic classifications defined in the British National Formulary (BNF) using the classification system prior to BNF edition 70. Each January the NHSBSA updates the classification of drugs within the BNF hierarchy which may involve some drugs changing classification between years of this publication. The categories of dependency-forming drugs do not align to specific sections of the BNF and have been filtered on chemical substance. A complete list can be found in appendix A of the background and methodology note that accompanies this release.

Many drugs have multiple uses, and although classified in the BNF by their primary therapeutic use may be issued to treat a condition outside of this. Due to this, these statistics may not give accurate estimations of prescribing to treat specific conditions.

These statistics do not exclude patients diagnosed with cancer and who are using an opioid to manage the pain that can be associated with malignant diseases, especially as part of end-of-life care. However, the PMR excluded opioids prescribed for cancer, using patient details from the PHE cancer registry. This means that some measures and the data for opioid pain medicines include more patients than the PMR analysis.

To exclude items used to treat an existing drug dependence or substance misuse disorder, drugs prescribed on FP10MDA instalment forms and from BNF Section 4.10 - Drugs used in substance dependence, were excluded from these statistics.

### 1.2. Key events {.toc-ignore}

```{r key events, results = TRUE}
Date  <- c("June 2015","September 2019","August 2021","April 2022","May 2022")
Event <- c("All Party Parliamentary Group for [Prescribed Drug Dependence](http://prescribeddrug.org/) launched to address the growing problem of prescribed drug dependence.", "Public Health England (PHE) [prescribed medicines review (PMR)](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/940255/PHE_PMR_report_Dec2020.pdf) intended to identify the scale, distribution and causes of prescription drug dependence published.", "Updated Opioid medicines and the risk of addiction [Safety leaflet](https://www.gov.uk/guidance/opioid-medicines-and-the-risk-of-addiction) published by Medicines and Healthcare products Regulatory Agency (MHRA)", "Medicines associated with dependence or withdrawal symptoms: safe prescribing and withdrawal management for adults guidance published by [NICE](https://www.nice.org.uk/guidance/ng215)","[Opioid comparator dashboard](https://www.nhsbsa.nhs.uk/access-our-data-products/epact2/dashboards-and-specifications/opioid-prescribing-comparators-dashboard) to support Primary Care Networks (PCN) and GP practices published by NHSBSA")
d <- data.frame(Date, Event)
knitr::kable(d, align = "ll")
```

### 1.3. Definitions {.toc-ignore}

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border(header = "Item", text = "A single unit of medication listed separately on a prescription form. In this publication, an example of an item would be Fluoxetine 20mg tables x56.", width = "100%")` 
::: 
::::

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("NIC", text = "The Net Ingredient Cost (NIC) is the basic price of the medication and the quantity prescribed. It does not include other fees incurred by dispensing contractors, such as controlled drug fees or the single activity fee. The basic price is determined by the [Drug Tariff](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff) or by the manufacturer, wholesaler, or supplier of the product.", width = "100%")`
::: 
::::

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Patient", text = "A unique NHS number captured from a prescription form or electronic prescription service (EPS) message.", width = "100%")`
::: 
::::

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Dependence", text = "The medicines included in these statistics are those that can cause issues with dependence. Dependence is an adaptation to repeated exposure to some drugs and medicines usually characterised by tolerance and withdrawal, though tolerance may not occur with some.", width = "100%")`
::: 
::::

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Tolerance", text = "Tolerance is a neuroadaptation arising from repeatedly taking some drugs and medicines, which can mean higher doses are required to achieve the same effect.", width = "100%")`
::: 
::::

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Addiction", text = "Addiction is the combination of dependence plus compulsive behaviours including patients not having control over doing, taking or using something to the point where it could be harmful to them.", width = "100%")`
::: 
::::

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Withdrawal", text = " Withdrawal is the side effects or physiological reactions that a patient experiences when they stop taking a medication.", width = "100%")`
::: 
::::

### 1.4. Patient identification {.toc-ignore}

When the NHSBSA processes prescriptions it is not always possible to capture the NHS number of the patient. Table 1 shows the proportion of items for which a patient could be identified. This means that the data relating to patient counts represents most, but not all, patients.

Due to an increase in digital prescription processing through the Electronic Prescription Service (EPS) during the COVID-19 pandemic, more patients were identified in 2020/21, 2021/22 and 2022/23 compared to previous years. As patient identification rates increased, any increases in the number of identified patients between periods are likely to be an overestimate of the actual increase in patient numbers. This is because the proportion of patients who could be identified has increased. Conversely, any decrease over the same period is likely to be an underestimate of the actual decrease.

Where patients are identified, their age is calculated on 30 September of the given financial year to assign them to a single age band. For patients where date of birth has not been captured, they have been included in an unknown category.

Gender information was not available from PDS for a small number of patients in each year, typically fewer than 100. This may be because it was not disclosed by the patient or not recorded by the organisation that collected the data.

These statistics do not include any information that is personally identifiable. You can find more information about how the NHSBSA protect personal information in the [confidentiality and access statement](https://www.nhsbsa.nhs.uk/policies-and-procedures).

#### **`r paste0("Table 1: The proportion of items for which an NHS number was recorded for listed Drug Categories per financial year")`**

```{r capture_rates}
knitr::kable(table_1, align = "lrrrrr")

get_download_button(title = "Download table data", data = table_1_data, filename = "table_1")
```

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

---

## 2. Results and commentary

### 2.1. Volume and cost

#### Number of dependency-forming medicine items prescribed and identified patients by financial year {.tabset}

<div class = "tabset">

##### Chart

###### Figure 1: Both items prescribed and identified patients have been declining since 2016/17

```{r}
figure_1

get_download_button(title = "Download chart data", data = figure_1_data, filename = "figure_1")
```

##### Table

###### Table 2: Both items prescribed and identified patients have been declining since 2016/17

```{r}
knitr::kable(table_2, align = "lrr")

get_download_button(title = "Download table data", data = figure_1_data, filename = "table_2")
```

</div>
  
Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>67 million items for drugs classed as dependency-forming prescribed in 2023/24.</b>", width = "100%")` 
::: 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>1% decrease in items from 2022/23 to 2023/24.</b>", width = "100%")` 
:::
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>7.1 million identified patients were prescribed a dependency-forming item in 2023/24.</b>", width = "100%")` 
:::
::::

There were 67 million items for drugs classed as dependency-forming prescribed in 2023/24. This was a 1% decrease from 2022/23. The number of items of dependency-forming medicines has been slowly decreasing since 2016/17.

The number of identified patients that received prescribing for a dependency-forming medication was 7.1 million in 2023/24. This was a 13% decrease from 8.1 million identified patients in 2015/16.

#### Cost of dependency-forming medicines per financial year {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 2: Costs have been declining since 2015/16, with the biggest decline between 2016/17 and 2018/19
  
```{r}
figure_2

get_download_button(title = "Download chart data", data = figure_2_data, filename = "figure_2")
```
##### Table

###### Table 3: Costs have been declining since 2015/16, with the biggest decline between 2016/17 and 2018/19

```{r}
knitr::kable(table_3, align = "lr")

get_download_button(title = "Download table data", data = figure_2_data, filename = "table_3")
```

</div>

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In 2023/24 the costs of prescribed drugs classed as dependency-forming was £370 million.</b>", width = "100%")` 
::: 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>There was a 3% decrease in costs from 2022/23 to 2023/24.</b>", width = "100%")` 
:::
::::

The costs for dependency-forming medicines was £370 million in 2023/24. This was a 52% decrease from £780 million in 2015/16, and a 3% decrease from £380 million in 2022/23.

The changes from 2016/17 to 2018/19 were due to pregabalin, a gabapenintoid coming off patent and entering Category M of the drug tariff, meaning cheaper generic equivalents could be prescribed from August 2017. This is more apparent in figure 5 and it's data.

#### Number of dependency-forming medicine items prescribed by Drug Category per financial year {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 3: Opioids are the most prescribed dependency-forming medicine
  
```{r}
figure_3

get_download_button(title = "Download chart data", data = figure_3_data, filename = "figure_3")
```

##### Table

###### Table 4: Opioids are the most prescribed dependency-forming medicine

```{r}
knitr::kable(table_4, align = "llr")

get_download_button(title = "Download table data", data = figure_3_data, filename = "table_4")
```

</div>

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>39 million opioid items prescribed in 2023/24.</b>", width = "100%")` 
::: 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>17 million gabapenintoid items prescribed in 2023/24.</b>", width = "100%")` 
:::
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>52% increase in gabapenintoid items from 2015/16 to 2023/24.</b>", width = "100%")` 
:::
::::

Opioid drugs remain the most prescribed dependency-forming medicine with 39 million items in 2023/24. This was a 7% decrease from 42 million items in 2015/16, and a 1% decrease from 39 million items in 2022/23. The number of opioid items prescribed has decreased every year since 2016/17.

Prescribing of gabapenintoids increased by 52% from 11 million items in 2015/16 to 17 million in 2023/24.

Benzodiazepine items decreased by 31% from 8.7 million items in 2015/16 to 6.1 million in 2023/24.

Z-drugs decreased by 20% from 6.4 million items in 2015/16 to 5.1 million in 2023/24.

#### Number of identified patients per 1,000 population by Drug Category per financial year {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 4: Opioids are the dependency-forming medicine with the highest identified patients per 1,000 population
  
```{r}
figure_4

get_download_button(title = "Download chart data", data = figure_4_data, filename = "figure_4")
```

##### Table

###### Table 5: Opioids are the dependency-forming medicine with the highest identified patients per 1,000 population

```{r}
knitr::kable(table_5, align = "llr")

get_download_button(title = "Download table data", data = figure_4_data, filename = "table_5")
```

</div>

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>96 patients per 1,000 population were prescribed an opioid drug in 2023/24.</b>", width = "100%")` 
::: 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>27 patients per 1,000 population were prescribed a gabapenintoid in 2023/24.</b>", width = "100%")` 
:::
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>17 patients per 1,000 population were prescribed a benzodiazepine in 2023/24.</b>", width = "100%")` 
:::
::::

In 2023/24 96 patients per 1,000 population were prescribed opioid drugs. This was an decease of 2 patients per 1,000 population from the 98 per 1,000 population in 2022/23.

Benzodiazepines decreased from 27 patients per 1,000 population in 2015/16 to 17 patients per 1,000 population in 2023/24. 

Z-drugs decreased from 19 patients per 1,000 population in 2015/16 to 13 patients per 1,000 population in 2023/24.

The patients per 1,000 population receiving gabapenintoids increased in 2023/24 to 27 patients per 1,000 population from 23 patients per 1,000 population in 2015/16.

The patients per 1,000 population are calculated using the [ONS population estimates](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates).

#### Cost of dependency-forming medicines prescribed by Drug Category per financial year {.tabset}

<div class = "tabset">

##### Chart

###### Figure 5: Opioids are the dependency-forming medicine with the highest costs

```{r}
figure_5

get_download_button(title = "Download chart data", data = figure_5_data, filename = "figure_5")
```

##### Table

###### Table 6: Opioids are the dependency-forming medicine with the highest costs

```{r}
knitr::kable(table_6, align = "llr")

get_download_button(title = "Download table data", data = figure_5_data, filename = "table_6")
```

</div>

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>Costs of opioid drugs prescribed in 2023/24 was £280 million.</b>", width = "100%")` 
::: 
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>2% decrease in opioid drug costs from 2022/23 to 2023/24.</b>", width = "100%")` 
:::
:::{.col-md-4} 
`r infoBox_no_border(header = "", text = "<b>12% decrease in benzodiapezine costs from 2022/23 to 2023/24.</b>", width = "100%")` 
:::
::::

In 2023/24, opioid drugs had a cost of £280 million. This was a decrease of 34% from £420 million in 2015/16, and an 2% decrease from 2022/23. These decreases in cost are much greater than the respective decreases in the number of prescribed items.

Gabapenintoids had a cost of £62 million in 2023/24, a decrease of 81% from the £320 million in 2015/16 and a 4% decrease from the £65 million in 2022/23. The costs of gabapenintoids decreased in consecutive years between 2017/18 and 2019/20 following pregabalin entering category M of the drug tariff meaning cheaper generic equivalents could be prescribed from August 2017.

Benzodiapezines decreased in cost by 12% from £31 million in 2022/23 to £27 million in 2023/24.

Z-drugs increased in cost by 32% from £4.1 million in 2022/23 to £5.5 million in 2023/24.

#### Average number of dependency-forming medicine items per patient by financial year {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 6: The average number of dependency-forming items per patient has increased between 2015/16 and 2023/24
  
```{r}
figure_6

get_download_button(title = "Download chart data", data = figure_6_data, filename = "figure_6")
```

##### Table

###### Table 7: The average number of dependency-forming items per patient has increased between 2015/16 and 2023/24

```{r}
knitr::kable(table_7, align = "lrrr")

get_download_button(title = "Download table data", data = figure_6_data, filename = "table_7")
```

</div>

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

The average number of items per patient has increased from 7.8 dependency-forming medicine items per patient in 2015/16 to 9.3 per patient in 2023/24.

This measure only includes prescribing of dependency-forming medicines and does not include any items prescribed from other BNF sections.

### 2.2. Patient demographics

#### Number of identified patients receiving dependency-forming medicine prescribing by gender and financial year {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 7: The overall split of male and female patients has remained consistent between 2015/16 and 2023/24
  
```{r}
figure_7

get_download_button(title = "Download chart data", data = figure_7_data, filename = "figure_7")
```

##### Table

###### Table 8: The overall split of male and female patients has remained consistent between 2015/16 and 2023/24

```{r}
knitr::kable(table_8, align = "llr")

get_download_button(title = "Download table data", data = figure_7_data, filename = "table_8")
```

</div>

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In 2023/24 61% of identified patients who were prescribed a dependency-forming item were female.</b>", width = "100%")` 
::: 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In 2023/24 39% of identified patients who were prescribed a dependency-forming item were male.</b>", width = "100%")` 
:::
::::

Although the total number of identified patients prescribed dependency-forming medicines has decreased between 2015/16 and 2023/24, the gender split remains consistent. In both 2015/16 and 2023/24, 61% of identified patients were female and 39% were male. However, in 2023/24, there were 580,000 fewer female identified patients and 450,000 fewer male identified patients compared to 2015/16.

Identified patients where their gender was unknown or indeterminate have been grouped together and can be found in the summary tables that accompany this release.

#### Number of identified patients receiving dependency-forming medicine prescribing by age and gender {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 8: Female patients aged 60 to 64 was the largest prescribing group for dependency-forming medicines in 2023/24
  
```{r}
figure_8

get_download_button(title = "Download chart data", data = figure_8_data, filename = "figure_8")
```

##### Table

###### Table 9: Female patients aged 60 to 64 was the largest prescribing group for dependency-forming medicines in 2023/24

```{r}
knitr::kable(table_9, align = "lllr")

get_download_button(title = "Download table data", data = figure_8_data, filename = "table_9")
```

</div>

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In 2023/24, 420,000 female patients aged 60 to 64 were prescribed a dependency-forming item.</b>", width = "100%")` 
::: 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In 2023/24, 300,000 male patients aged 60 to 64 were prescribed a dependency-forming item.</b>", width = "100%")` 
:::
::::

Prescribing of dependency forming medicines peaks for both females and males age 60 to 64 years. In 2023/24, they made up 10% of all patients who received a dependency-forming item.

#### Number of identified patients receiving dependency-forming medicines prescribing by IMD quintile {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 9: More people were prescribed dependency-forming medicines in more deprived areas
  
```{r}
figure_9

get_download_button(title = "Download chart data", data = figure_9_data, filename = "figure_9")
```

##### Table

###### Table 10: More people were prescribed dependency-forming medicines in more deprived areas

```{r}
knitr::kable(table_10, align = "llr")

get_download_button(title = "Download table data", data = figure_9_data, filename = "table_10")
```

</div>

Source: [Dependency-forming medicines summary tables - Costs and items](`r config$costs_items_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In 2023/24, 1.8 million patients who were prescribed a dependency-forming item were from the most deprived areas in England.</b>", width = "100%")` 
::: 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In 2023/24, 1.2 million patients who were prescribed a dependency-forming item were from the least deprived areas in England.</b>", width = "100%")` 
:::
::::

In 2023/24, 1.8 million identified patients in England's most deprived areas were prescribed dependency-forming medicines, 57% more than the 1.2 million in the least deprived areas. This trend, with higher identified patient numbers in more deprived areas, has been consistent since 2015/16.

### 2.3. Co-prescribing of drug categories

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("", text = "Co-prescribing is where a patient is receiving drugs from more than one category of dependency-forming medicine. Antidepressants which have been excluded from the previous sections have been included here because of the increased risk factors when combined with dependency-forming medicines. It is not possible to distinguish whether multiple prescriptions which have been reported for the same month were given consecutively or concurrently. As such, some activity will show as co-prescribing when in fact the individual was prescribed one medicine and another separately, and both were reported in the same month. March 2024 has been used for this analysis as the most recent month of available data and as it was representative of the recent trends in co-prescribing.", width = "100%")`
:::
::::

#### Number of identified patients receiving more than one category of dependency-forming medicines prescribing by number of categories {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 10: Almost a quarter of patients who received a prescription for dependency-forming medicines were prescribed drugs from more than one category in March 2024
  
```{r}
figure_10

get_download_button(title = "Download chart data", data = figure_10_data, filename = "figure_10")
```

##### Table

###### Table 11: Almost a quarter of patients who received a prescription for dependency-forming medicines were prescribed drugs from more than one category in March 2024

```{r}
knitr::kable(table_11, align = "lrr")

get_download_button(title = "Download table data", data = figure_10_data, filename = "table_11")
```

</div>

Source: [Dependency-forming medicines summary tables - Co-prescribing](`r config$co_prescribing_excel`)

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In March 2024, 24% of patients who received a prescription for dependency-forming medicines were prescribed drugs from more than one category.</b>", width = "100%")` 
::: 
:::{.col-md-6} 
`r infoBox_no_border(header = "", text = "<b>In March 2024, less than 0.1% of patients who received a prescription for dependency-forming medicines were prescribed drugs from all 5 categories.</b>", width = "100%")` 
:::
::::

In total, an estimated 6.5 million identified patients received a prescription in March 2024 in at least one of the categories. Of these, 1.2 million were receiving prescriptions in 2 categories, representing 18% of patients. Less than 0.1% of patients received a drug from all 5 categories of dependency-forming medication. 

#### Number of identified patients receiving prescribing of a combination of two dependency-forming medicines {.tabset}

<div class = "tabset">
  
##### Chart
  
###### Figure 11: Opioids and antidepressants was the most popular combination of drugs for those patients receiving items from 2 categories of dependency-forming medicines in March 2024
  
```{r}
figure_11

get_download_button(title = "Download chart data", data = figure_11_data, filename = "figure_11")
```

##### Table

###### Table 12: Opioids and antidepressants was the most popular combination of drugs for those patients receiving items from 2 categories of dependency-forming medicines in March 2024

```{r}
knitr::kable(table_12, align = "llr")

get_download_button(title = "Download table data", data = figure_11_data, filename = "table_12")
```

</div>

Source: [Dependency-forming medicines summary tables - Co-prescribing](`r config$co_prescribing_excel`)

In March 2024, the most popular combination of drugs for those patients receiving items from 2 categories of dependency-forming medicines was opioids and antidepressants with 620,000 patients. This was 52% of patients who received prescribing in two categories of dependency-forming medicines. 

This was followed by gabapentinoids and antidepressants with 220,000 patients. This was 19% of patients who received prescribing in 2 categories of dependency-forming medicines.

---

## 3. Background

### 3.1. Opioid pain medicine {.toc-ignore}

Opioids provide pain relief by acting on areas in the spinal cord and brain to block the transmission of pain signals.

Most opioids are schedule 2 controlled drugs, unless very low strength which may be schedule 5, and are available in a wide variety of medication forms.

Opioids should only be considered for the short-to-medium-term treatment of chronic non-malignant pain, when other therapies have been insufficient and the benefits of use are considered to outweigh the risks of harm.

Opioid analgesics are usually used for palliative care, where potential for dependence is not a deterrent, and chronic (lasting more than 12 weeks) moderate-to-severe pain exists where other treatments have been insufficient due to the potential for dependence.

You can find out [more information on opioid analgesics](https://bnf.nice.org.uk/treatment-summaries/analgesics/#opioid-analgesics-and-dependence) on the NICE website. [Resources for the prescribing of opioids](https://fpm.ac.uk/opioids-aware) have been produced by the Faculty of Pain Medicine in partnership with PHE.

### 3.2. Gabapentinoids {.toc-ignore}

Gabapentinoids is the combined name for gabapentin and pregabalin which are antiepileptic drugs also used in the treatment of neuropathic pain and in the case of pregabalin, anxiety.

In epilepsy, gabapentinoids stop seizures by reducing the abnormal electrical activity in the brain.

With nerve pain, they block pain by affecting the pain messages travelling through the brain and down the spine.

When pregabalin is used to treat anxiety, it prevents the brain from releasing the chemicals that cause anxiety.

Both gabapenintoids are schedule 3 controlled drugs and are available as capsules, tablets, or a liquid.

You can find out more about [gabapentin](https://www.nhs.uk/medicines/gabapentin/) and [pregabalin](https://www.nhs.uk/medicines/pregabalin/) on the NHS website.

### 3.3. Benzodiazepines {.toc-ignore}

Benzodiazepines are a commonly used hypnotic and anxiolytic medicine. Hypnotics and anxiolytics are used to treat insomnia and anxiety respectively. Benzodiazepines work by increasing the effects of a calming chemical in the brain called gamma-aminobutyric acid (GABA).

Benzodiazepines are indicated for the short-term relief of severe anxiety. Long-term use should be avoided and should also only be used to treat insomnia only when it is severe, disabling, or causing the patient extreme distress

The majority of benzodiazepines are schedule 4 controlled drugs with some belonging to schedule 3, and are available as capsules, tablets, injectables, suppositories or a liquid.

Insomnia is difficulty getting to sleep or staying asleep for long enough to feel refreshed in the morning, despite there being enough opportunity to sleep. An insomniac may also experience:

* waking in the night
* not feeling refreshed after sleep and not being able to function normally during the day
* feeling irritable and tired and finding it difficult to concentrate
* waking when they have been disturbed from sleep by pain or noise
* waking early in the morning

Anxiety is a feeling of unease, such as worry or fear, which can be mild or severe. Everyone experiences feelings of anxiety at some point in their life and feeling anxious is sometimes perfectly normal. However, people with generalised anxiety disorder (GAD) find it hard to control their worries. Their feelings of anxiety are more constant and often affect their daily life. There are several conditions for which anxiety is the main symptom. Panic disorder, phobias and post-traumatic stress disorder can all cause severe anxiety.

You can find more information about [insomnia](https://www.nhs.uk/conditions/insomnia/) and [anxiety](https://www.nhs.uk/conditions/generalised-anxiety-disorder/) from the NHS website, and further information about [hypnotics and anxiolytics](https://bnf.nice.org.uk/treatment-summaries/hypnotics-and-anxiolytics/) at the NICE website, though this includes drugs other than benzodiazepines.

### 3.4. Z-drugs {.toc-ignore}

Z drugs are are non-benzodiazepine hypnotics made up of zaleplon, zolpidem and zopiclone. As hypnotics they are also used to treat insomnia. Z drugs work by affecting a calming chemical in the brain called gamma-aminobutyric acid (GABA).

Zolpidem is a schedule 4 controlled drug and is available as tablets or a powder. Zaleplon is not a controlled drug and is available as a capsule and zopiclone is not a controlled drug and is available as capsules, tablets, or a liquid.

You can find more information about [insomnia](https://www.nhs.uk/conditions/insomnia/) on the NHS website and further information about [hypnotics and anxiolytics](https://bnf.nice.org.uk/treatment-summaries/hypnotics-and-anxiolytics/) at the NICE website.

### 3.5. Antidepressants {.toc-ignore}

Antidepressant drugs are licensed to treat major depression. Health professionals use the words depression, depressive illness or clinical depression to refer to depression. It is a serious illness and very different from the common experience of feeling unhappy or fed up for a short period of time. Depressed people may have feelings of extreme sadness that can last for a long time. These feelings are severe enough to interfere with daily life, and can last for weeks, months or years, rather than days.

It should be noted that antidepressant drugs are used for indications other than depression. For example, they can used for migraine, chronic pain, Myalgic Encephalomyelitis (ME), or a range of other conditions. Clinical indication is not captured by the NHSBSA. Therefore, the statistics on these drugs do not relate solely to prescribing for depression.

You can find more [information about depression](https://www.nhs.uk/conditions/clinical-depression/) on the NHS website.

---

## 4. About these statistics

Further information on the methodology used in this publication and further background information is available in our [Background Information and Methodology](`r config$background_link`) supporting document.

### 4.1. Patient counts {.toc-ignore}

The patient counts shown in these statistics should only be analysed at the level at which they are presented. Adding together any patient counts is likely to result in an overestimate of the number of patients. A person will be included, or counted, in each category or time period in which they received relevant prescriptions. For example, if a patient received a prescription item for a dependency-forming medicine product in 2018/19 and another in 2019/20, then adding together those totals would count that patient twice. For the same reason, data on patient counts for different drug categories should not be added together.

### 4.2. Patient age and gender {.toc-ignore}

The age and gender of patients used in these statistics is derived from data provided by the NHS Personal Demographics Service (PDS) for NHS numbers that have been successfully verified by them. A patient’s age, used to assign them to an age group, has been calculated on 30 September for the given financial year. It is possible that a patient’s PDS information may change over the course of the year, in these cases patients may be subject to multiple counting in these analyses.

### 4.3. Index of deprivation {.toc-ignore}

The English Indices of Deprivation 2019 have been used to provide a measure of patient deprivation. The English Indices of Deprivation are an official national measure of deprivation that follows an established methodological framework to capture a wide range of individuals living conditions.

The reported IMD quintile, is derived from the postcode of the patient an item has been prescribed to. Quintile 1 represents the 20% most deprived areas and quintile 5 is the 20% least deprived areas. There are a small number of items each year that we have reported as having an unknown IMD quintile. These are items where we have been unable to match the patient postcode to a postcode in the National Statistics Postcode Lookup (NSPL) - May 2023.

IMD deciles are calculated by ranking census lower-layer super output areas (LSOA) from most deprived to least deprived and dividing them into 10 equal groups. These range from the most deprived 10% (decile 1) of small areas nationally to the least deprived 10% (decile 10) of small areas nationally. We have aggregated these deciles into quintiles in this publication, for use alongside the [NHS Core20PLUS5 approach](https://www.england.nhs.uk/about/equality/equality-hub/national-healthcare-inequalities-improvement-programme/core20plus5/).

### 4.4. Geographies included in this publication {.toc-ignore}

The patient deprivation measures given in these statistics are based upon the LSOA of the postcode of the patient, as matched to the May 2023 NSPL file. However, higher geographies included in the statistical summary tables of this publication, such as ICB, use NHSBSA administrative records, not geographical boundaries, and more closely reflect the operational organisation of practices than other geographical data sources.

### 4.5. Co-prescribing measures {.toc-ignore}

In these statistics, co-prescribing refers to the reporting of 2 or more drug categories of medicine for the same identified patient in the same month. Co-prescribing is reported based on the number of categories, up to 5, that were reported in the same month, with greater than one deemed to be co-prescribing. The main limitation is that it is not possible to distinguish whether multiple prescriptions that have been reported for the same month were given consecutively or concurrently. As such, some activity will be flagged in this analysis as co-prescribing, implying they were received at the same time, when in fact the individual was prescribed one medicine and then another separately, and both were reported in the same month.

---

## 5. Rounding

The high-level figures in this statistical summary have been rounded as per the table below:

```{r rounding_table}
rounding_table <- data.frame(
  "From" = c("0", "1,001", "10,001", "100,001", "1,000,001", "10,000,001", "100,000,001"),
  "To" = c("1,000", "10,000", "100,000", "1,000,000", "10,000,000", "100,000,000", "100,000,000,000"),
  "Round to nearest" = c("1", "100", "1,000", "10,000", "100,000", "1,000,000", "10,000,000"),
  check.names = FALSE
)

knitr::kable(rounding_table, align = "rrr")
```

All changes and totals are calculated prior to rounding. Percentage changes are calculated prior to rounding and then are rounded to the nearest whole number. As all figures within this statistical summary have been rounded, they may not match totals elsewhere when aggregated.

The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

---

## 6. Statistical disclosure control

Statistical disclosure control has been applied to these statistics. Patient count, items, and net ingredient cost (NIC) have been redacted in the supporting summary tables if they relate to fewer than 5 patients. [Further information about our statistical disclosure control protocol can be found on our website](https://www.nhsbsa.nhs.uk/policies-and-procedures).

---

## 7. Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: accessibility@nhsbsa.nhs.uk

**Telephone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We’ll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This email address is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Feedback and contact us’ section of this page.

Read our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---

## 8. Feedback and contact us

Feedback is important to us. We welcome any questions and comments relating to these statistics.

You can complete a [short survey about this publication](`r config$stats_survey_link`) to help us improve the Official Statistics that we produce. All responses will remain anonymous, and individuals will not be identifiable in any report that we produce.

You can read our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to find out how your data is used and stored.

You can contact us by:

**Email:** statistics@nhsbsa.nhs.uk

**You can also write to us at:**

NHSBSA - Statistics<br>
NHS Business Services Authority<br>
Stella House<br>
Goldcrest Way<br>
Newburn Riverside<br>
Newcastle upon Tyne<br>
NE15 8NY

**Responsible statistician:** `r config$responsible_statistician`

</main>
